---
title: Designing PCR Diagnostics that Discriminate HPV Serotypes by DNA Sequence
author: chebuu
bibliography: citations.bib
output: pdf_document
---


<!-- --- -->
<!-- ... -->
<!-- includes: -->
<!--   in_header:  -->
<!--     - preamble.tex -->
<!--     - common.tex -->
<!-- ... -->
<!-- --- -->


```{r setup, include=F}
knitr::opts_chunk$set(echo = T)
```

# Introduction
The first infectious cause of cancer was identified in 1911 by Peyton Rous who demonstrated the transmissibility of a tumor in fowl through injection of sub 0.2 micron filtrate of the tumor \cite{Rous_1911}.

#### Installing OligAarrayAux

Download [OligoArrayAux](http://unafold.rna.albany.edu/?q=DINAMelt/OligoArrayAux) and see the following vignette for installation instructions.

```{r oaux, eval=F}
library(devtools)

devtools::install_github('chebuu/Design-Group-Specific-Primers')

library(Design_Group_Specific_Primers)

vignette('Installing-OligoArrayAux', package = 'Design_Group_Specific_Primers')
```

### Install Bioconductor packages
```{r biocs, eval=F, error=F, warning=F, message=F}
# Install from Bioconductor
if (!requireNamespace("BiocManager", quietly = T))
  install.packages("BiocManager")
BiocManager::install("Biostrings")
BiocManager::install("DECIPHER")

# Open Biostrings vignettes
library(Biostrings)
browseVignettes("Biostrings")

# Open DECIPHER docs and vignettes
library(DECIPHER)
help("DECIPHER")
browseVignettes("DECIPHER")
```

```{}
# https://rdrr.io/cran/dpcR/f/vignettes/overview.Rmd

# https://bioconductor.org/packages/release/bioc/html/twoddpcr.html
# https://bioconductor.org/packages/release/bioc/html/ddCt.html
# https://bioconductor.org/packages/3.5/bioc/html/mBPCR.html
```
# PaVE Data
[https://pave.niaid.nih.gov](https://pave.niaid.nih.gov)
```{}
https://pave.niaid.nih.gov/#search/search_database/kw
?dbNamespace=Genomes
&includeNR=true
&refCloneOnly=false
&sort=Locus_ID
&sortType=true
&page=600&start=1
&text=Human&showTable=1
&
```

# NCBI Data

```{r packs, eval=T, error=F, warning=F, message=F}
library(dplyr)
library(tidyr)
library(stringr)

library(Biostrings)
library(DECIPHER)
```

I included GenBank queries for all the HPV DNA in this repo. If you want to run them just paste in the search bar at: [https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore) 

## All HPV
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
 "Human papillomavirus"[Primary Organism] 
 AND viruses[filter] 
 NOT Polyamides[All Fields] 
 NOT Method[All Fields] 
 NOT Patent[All Fields]
```

## Complete Genomes
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
"Human papillomavirus"[Primary Organism] 
AND "complete genome"[All Fields]
NOT Isolate[Title] 
```

## Oral Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
https://www.ncbi.nlm.nih.gov/nuccore
     "Human papillomavirus"[Primary Organism] 
 AND viruses[filter] 
 NOT Polyamides[All Fields] 
 NOT Method[All Fields] 
 NOT Patent[All Fields] 
 AND Oral[All Fields] 
```

### L1 gene
A collection of *l1* DNA sequences (avg. ~600bp) from various oral isolates of HPV (extracted from results of GenBank query above) are aligned and loaded in SQLite (RAM) along with annotations from GenBank that describe the serotype of each isolate.

```{r oral_L1}
doAlignment <- function(xset) {
    useqs <- unique(xset)
    uidxs <- match(xset, useqs)
    aseqs <- AlignSeqs(useqs, verbose=F) 
    aseqs[uidxs]
}

dbConn <- dbConnect(SQLite(), ':memory:')

Oral.L1.seqs <- readDNAStringSet('HPV.oral.L1.fasta') %>% doAlignment
Oral.L1.vars <- read.csv('HPV.oral.L1.csv', stringsAsFactors = F)

print(Oral.L1.seqs)

Seqs2DB(Oral.L1.seqs, 'XStringSet', dbConn, '')
Add2DB(Oral.L1.vars %>% mutate(identifier = SVAR), dbConn)

dbGetQuery(dbConn, "select * from Seqs") %>% head(4)
```

Design HPV16-specific F/R primers using `DECIPHER` utilities.

```{r oral_l1_olg, include=T}
tiles.L1 <- TileSeqs(
  dbFile = dbConn,
  minLength = 18,
  maxLength = 29,
  minCoverage = 0.8
)

print(
  tiles.L1[,c(6,11)] %>% sample_n(6)
)

oligos.L1 <- DesignPrimers(
  tiles = tiles.L1,
  identifier = 'HPV16',
  worstScore = -1E3,
  maxPermutations = 5,
  minGroupCoverage = 0.85,
  minCoverage = 0.85,
  minLength = 20,
  maxLength = 28
)

head(oligos.L1)
```

Honestly, I don't really understand the dimensions of the dataframe output by `DECIPHER::DesignPimers`. If I remember correctly, it's a 7x17 matrix where each cell holds a list of 5 oligos. Unfortunately, I can't print the matrix to stdout, which makes debugging slow, so I'll just deal with this issue later I guess...

Plot hybridization curves for the reverse compliment of each primer pair (5) in each primer set (7). 

```{r melt_curves, include=T}
plotMeltCurves <- function(temps, effs) {
  plot(
    temps, effs[,1], ylim=c(0,1),
    ylab="Hybridization Efficiency",
    xlab=expression(paste("Temperature (", degree, "C)", sep="")),
    type="l", lwd=2, col="Blue", main="Denaturation Plot"
  )
  lines(temps, effs[,2], col="Red", lwd=2)
  abline(h=0.5, lty=2, lwd=2, col="Orange")
  abline(v=64, lty=2, lwd=2, col="Green")
  legend(
    "topright",
    legend = c(
      "Forward Primer",
      "Reverse Primer",
      "50% Efficiency",
      "Annealing Temperature"
    ),
    col = c("Blue", "Red", "Orange", "Green"),
    lwd = c(2, 2, 2, 2), lty = c(1, 1, 2, 2)
  )
}

meltCurves <- 
  function(primers, target=reverseComplement(DNAStringSet(primers)), temps=60:75, P=4e-7, ions=.225, doPlot=TRUE, ...) {
    fxn <- function(temp) CalculateEfficiencyPCR(primers, target, temp, P=P, ions=ions, ...)
    effs <- matrix(unlist(lapply(temps, fxn)), ncol=2, byrow=TRUE)
    if (doPlot) plotMeltCurves(temps, effs)
    list(temps = temps, effs = effs)
  }

nsets <- nrow(oligos.L1)
npair <- ncol(oligos.L1)

for (i in 1:nsets)
  for (j in 1:npair)
    tryCatch(
      {
        c(
          oligos.L1$forward_primer[i,j],
          oligos.L1$reverse_primer[i,j]
        ) %>% meltCurves
      },
      error = function(e) NULL
        # # ¿ --- It's supposed to be 7o x 17p --- ?
        # warning(sprintf('%s [iteration i=%s j=%s]', e, i, j))
    )

```

More primer designs (RFLP, sequencing, etc.):

```{r oral_l1_seq, include=T}
TYPE <- 'sequence'
MIN_LENGTH <- 15
MAX_LENGTH <- 25
MIN_SIZE <- 60
MAX_SIZE <- 100
LEVELS <- 2
RESOLUTION <- 3
ENZYMES <- NULL

DesignSignatures(
  dbConn,
  type=TYPE,
  minLength=MIN_LENGTH,
  maxLength=MAX_LENGTH,
  minProductSize=MIN_SIZE,
  maxProductSize=MAX_SIZE,
  resolution=RESOLUTION,
  levels=LEVELS,
  enzymes=ENZYMES
)
```

```{r oral_l1_rflp}
data(RESTRICTION_ENZYMES)

TYPE <- 'melt'
LEVELS <- 4
MIN_LENGTH <- 15
MAX_LENGTH <- 25
MIN_SIZE <- 60
MAX_SIZE <- 200
RESOLUTION <- seq(75, 300, 15)

lapply(
  rep(sample(1:3, 3), 3),  # Random RE combinations
  function(i)
    DesignSignatures(
      dbConn,
      type=TYPE,
      enzymes=RESTRICTION_ENZYMES[i],
      minProductSize=MIN_SIZE,
      maxProductSize=MAX_SIZE,
      resolution=RESOLUTION, 
      levels=LEVELS
    )
)

# TYPE <- 'length'
# LEVELS <- 2
# MIN_SIZE <- 200
# MAX_SIZE <- 1400
# RESOLUTION <- c(
#   seq(200, 700, 3),
#   seq(705, 1000, 5),
#   seq(1010, 1400, 10)
# )
# 
# TYPE <- 'melt'
# MIN_SIZE <- 55
# MAX_SIZE <- 400
```

## Anal Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
 "Human papillomavirus"[Primary Organism]
 AND viruses[filter] 
 NOT Polyamides[All Fields] 
 NOT Method[All Fields] 
 NOT Patent[All Fields]
 AND Anal[All Fields]
```

## Cutaneous Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
"Human papillomavirus"[Primary Organism]
NOT Polyamides[All Fields] 
NOT Method[All Fields] 
NOT Patent[All Fields]
AND Anal[All Fields]
AND "Complete Genome"[All Fields]
```

```{r disconnect, eval=F}
tryCatch({dbDisconnect(dbconn)}, error=warning)
tryCatch({dbDisconnect(dbConn)}, error=warning)
```
