---
title: Designing PCR Diagnostics that Discriminate HPV Serotypes by DNA Sequence
author: chebuu
bibliography: citations.bib
output: pdf_document
---


<!-- --- -->
<!-- ... -->
<!-- includes: -->
<!--   in_header:  -->
<!--     - preamble.tex -->
<!--     - common.tex -->
<!-- ... -->
<!-- --- -->


```{r setup, include=F}
library(knitr)
knitr::opts_chunk$set(eval = T, echo = T)
```

# Introduction
The first infectious cause of cancer was identified in 1911 by Peyton Rous who demonstrated the transmissibility of a tumor in fowl through injection of sub 0.2 micron filtrate of the tumor \cite{Rous_1911}.

#### Installing this package
```{r hpvs}
devtools::install_github('Chebuu/HPV-Serovars')
library(HPVSerovars)
```

(Not run) This vignette is packaged and accessible via: `browseVignettes('HPVSerovars')`.

#### Installing OligAarrayAux

Download [OligoArrayAux](http://unafold.rna.albany.edu/?q=DINAMelt/OligoArrayAux) and see the following vignette for installation instructions.

```{r oaux, eval=F}
library(devtools)

devtools::install_github('chebuu/Design-Group-Specific-Primers')

library(Design_Group_Specific_Primers)

vignette('Installing-OligoArrayAux', package = 'Design_Group_Specific_Primers')
```

#### Installing OpenPrimer Dependencies

From the [openPimer vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/openPrimeR/inst/doc/openPrimeR_vignette.html):

openPrimeR requires external programs for some features, particularly for computing the physicochemical properties of primers. Please make sure you have the following tools installed on your system such that they are in your system’s path:

[MELTING](http://www.ebi.ac.uk/biomodels/tools/melting/) (>= 5.1.1): For melting temperature computations.
[ViennaRNA](http://www.tbi.univie.ac.at/RNA/) (>= 2.2.4): For secondary structure prediction.
[OligoArrayAux](http://unafold.rna.albany.edu/OligoArrayAux.php) (>= 3.8): For primer efficiency computations as performed by DECIPHER.
[MAFFT](http://mafft.cbrc.jp/alignment/software/) (>= 7.305): For computing multiple sequence alignments.
[Pandoc](http://pandoc.org/) (>= 1.19.1): For creating PDF reports.

If you would like to be able to access the immunoglobulin repository IMGT from the openPrimeR Shiny app, you should additionally fulfill the following dependencies:

[PhantomJS](http://phantomjs.org/) (>= 2.1): For headless website calls.
[Python](http://www.python.org/) (>=2.7.9) and the [selenium](http://selenium-python.readthedocs.io/) (>=3.0.1) module: For data extraction scripts.
openPrimeR will automatically check for all dependencies and inform you about any missing dependencies when the package is attached:

```{r openprimer}
library(openPrimeR)
```

Note that the tool is still functional if there are missing external programs. However, we recommend that all dependencies are fulfilled to guarantee the best user experience.

### Install Bioconductor packages
[https://www.bioconductor.org/](https://www.bioconductor.org/)
```{r biocs, eval=F, error=F, warning=F, message=F}
# Install from Bioconductor
if (!requireNamespace("BiocManager", quietly = T))
  install.packages("BiocManager")
BiocManager::install("Biostrings")
BiocManager::install("DECIPHER")

# Open Biostrings vignettes
library(Biostrings)
browseVignettes("Biostrings")

# Open DECIPHER docs and vignettes
library(DECIPHER)
help("DECIPHER")
browseVignettes("DECIPHER")
```

#### Make alignment utils:

```{r alignment}
doAlignment <- function(xset) {
  useqs <- unique(xset)
  uidxs <- match(xset, useqs)
  aseqs <- AlignSeqs(useqs, verbose=F) 
  aseqs[uidxs]
}
```

#### Quick-ref
Some references to remind me of packages, methods, etc.:
```{}
# CRAN
# https://rdrr.io/cran/dpcR/f/vignettes/overview.Rmd

# BIOC
# https://bioconductor.org/packages/release/bioc/html/twoddpcr.html
# https://bioconductor.org/packages/release/bioc/html/ddCt.html
# https://bioconductor.org/packages/3.5/bioc/html/mBPCR.html

# Other
# https://doi.org/10.1007/s13402-017-0331-y
```

# Wiki Pheno Data
I copied the tables from [Wiki: Human_papillomavirus_infection ](https://en.wikipedia.org/wiki/Human_papillomavirus_infection) and included them in the R package assoc. w/ this document `github.com/Chebuu/HPV-Serovars`.
```{r wikidata}
data(HPV.Sequelae)
data(HPV.GenitalCancerRisk)
```

The Wiki data contains phenotypic categorizations of several HPV strains. It's useful for testing.

```{r wikiprint}
HPV <- c(HPV.Sequelae, HPV.GenitalCancerRisk)
p.1 <- unique(unlist(HPV))

knitr::kable(
  sprintf('HPV-%s', sort(p.1)), col.names = '',
  caption = sprintf(
    'Unique HPV serotypes (N=%s) (Total=%s): ', 
    length(p.1), length(HPV)
  )
)

hist(unlist(HPV), max(unlist(HPV)), col='blue'); head(HPV, 4)
```


# PaVE Data

Reference genomes for all serovars listed in `data(HPV.Sequelae)`and `data(HPV.GenitalCancerRisk)`were selected from among the results of the following [PaVEPaVE](https://pave.niaid.nih.gov) query:

```{} 
https://pave.niaid.nih.gov/#search/search_database/kw
?dbNamespace=Genomes
&includeNR=true
&refCloneOnly=false
&sort=Locus_ID
&sortType=true
&page=600&start=1
&text=Human&showTable=1
&
```

```{r}

library(dplyr)
library(tidyr)
library(stringr)

library(Biostrings)
library(DECIPHER)

# data(PaVE.Wikiv)
data(HPV.REF.PaVE.seqs)
data(HPV.REF.PaVE.vars)

testVars <- lapply(grep('[1-9]*', HPV.REF.PaVE.vars, value = T), function(x) {
  sapply(HPV.Sequelae, function(y) any(x %in% y))
})

tryCatch(dbDisconnect(dbConn), error = warning)
dbConn <- dbConnect(SQLite(), ':memory:')
DECIPHER::Seqs2DB(HPV.REF.PaVE.seqs, 'XStringSet', dbConn, '')
DECIPHER::Add2DB(
  data.frame(
    # identifier = testVars
    identifier = lapply(HPV.REF.PaVE.vars, function(x) grep('[1-9]', x, perl = T))
  ), 
  dbConn, verbose = F
)

# # DEL
# library(Biostrings)
# PaVE.Wikiv<-Biostrings::readDNAStringSet('../inst/extdata/HPV.REFS.PaVE.wikivars.fasta')
# usethis::use_data(PaVE.Wikiv)
#
# https://www.youtube.com/watch?v=zu8Z9-kebKQ&list=PLjiXAZO27elDHGlQwfd06r7coiFtpPkvI&index=5
#
# # END
```

# NCBI Data

```{r packs, error=F, warning=F, message=F}
library(dplyr)
library(tidyr)
library(stringr)

library(Biostrings)
library(DECIPHER)
```

I included GenBank queries for all the HPV DNA in this repo. If you want to run them just paste in the search bar at: [https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore) 

## All HPV
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
 "Human papillomavirus"[Primary Organism] 
 AND viruses[filter] 
 NOT Polyamides[All Fields] 
 NOT Method[All Fields] 
 NOT Patent[All Fields]
```

## Complete Genomes
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
"Human papillomavirus"[Primary Organism] 
AND "complete genome"[All Fields]
NOT Isolate[Title] 
```

## Oral Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
"Human papillomavirus"[Primary Organism] 
AND viruses[filter] 
NOT Polyamides[All Fields] 
NOT Method[All Fields] 
NOT Patent[All Fields] 
AND Oral[All Fields] 
```

### L1 gene
A collection of *l1* DNA sequences (avg. ~600bp) from various oral isolates of HPV (extracted from results of GenBank query above) are aligned and loaded in SQLite (RAM) along with annotations from GenBank that describe the serotype of each isolate.

```{r oral_L1}
dbConn <- dbConnect(SQLite(), ':memory:')

Oral.L1.seqs <- readDNAStringSet('HPV.oral.L1.fasta') %>% doAlignment
Oral.L1.vars <- read.csv('HPV.oral.L1.csv', stringsAsFactors = F)

print(Oral.L1.seqs)

Seqs2DB(Oral.L1.seqs, 'XStringSet', dbConn, '')
Add2DB(Oral.L1.vars %>% mutate(identifier = SVAR), dbConn)

dbGetQuery(dbConn, "select * from Seqs") %>% head(4)
```

Design HPV16-specific F/R primers using `DECIPHER` utilities.

```{r oral_l1_olg}
tiles.L1 <- TileSeqs(
  dbFile = dbConn,
  minLength = 18,
  maxLength = 29,
  minCoverage = 0.8
)

print(
  tiles.L1[,c(6,11)] %>% sample_n(6)
)

oligos.L1 <- DesignPrimers(
  tiles = tiles.L1,
  identifier = 'HPV16',
  worstScore = -1E3,
  maxPermutations = 5,
  minGroupCoverage = 0.85,
  minCoverage = 0.85,
  minLength = 20,
  maxLength = 28
)

head(oligos.L1)
```

Honestly, I don't really understand the dimensions of the dataframe output by `DECIPHER::DesignPimers`. If I remember correctly, it's a 7x17 matrix where each cell holds a list of 5 oligos. Unfortunately, I can't print the matrix to stdout, which makes debugging slow, so I'll just deal with this issue later I guess...

Plot hybridization curves for the reverse compliment of each primer pair (5) in each primer set (7). 

```{r melt_curves, include=T}
plotMeltCurves <- function(temps, effs) {
  plot(
    temps, effs[,1], ylim=c(0,1),
    ylab="Hybridization Efficiency",
    xlab=expression(paste("Temperature (", degree, "C)", sep="")),
    type="l", lwd=2, col="Blue", main="Denaturation Plot"
  )
  lines(temps, effs[,2], col="Red", lwd=2)
  abline(h=0.5, lty=2, lwd=2, col="Orange")
  abline(v=64, lty=2, lwd=2, col="Green")
  legend(
    "topright",
    legend = c(
      "Forward Primer",
      "Reverse Primer",
      "50% Efficiency",
      "Annealing Temperature"
    ),
    col = c("Blue", "Red", "Orange", "Green"),
    lwd = c(2, 2, 2, 2), lty = c(1, 1, 2, 2)
  )
}

meltCurves <- 
  function(primers, target=reverseComplement(DNAStringSet(primers)), temps=60:75, P=4e-7, ions=.225, doPlot=TRUE, ...) {
    fxn <- function(temp) CalculateEfficiencyPCR(primers, target, temp, P=P, ions=ions, ...)
    effs <- matrix(unlist(lapply(temps, fxn)), ncol=2, byrow=TRUE)
    if (doPlot) plotMeltCurves(temps, effs)
    list(temps = temps, effs = effs)
  }

nsets <- nrow(oligos.L1)
npair <- ncol(oligos.L1)

for (i in 1:nsets)
  for (j in 1:npair)
    tryCatch(
      {
        c(
          oligos.L1$forward_primer[i,j],
          oligos.L1$reverse_primer[i,j]
        ) %>% meltCurves
      },
      error = function(e) NULL
        # # ¿ --- It's supposed to be 7o x 17p --- ?
        # warning(sprintf('%s [iteration i=%s j=%s]', e, i, j))
    )

```

More primer designs (RFLP, sequencing, etc.):

```{r oral_l1_seq, include=T}
TYPE <- 'sequence'
MIN_LENGTH <- 15
MAX_LENGTH <- 25
MIN_SIZE <- 60
MAX_SIZE <- 100
LEVELS <- 2
RESOLUTION <- 3
ENZYMES <- NULL

DesignSignatures(
  dbConn,
  type = TYPE,
  minLength = MIN_LENGTH,
  maxLength = MAX_LENGTH,
  minProductSize = MIN_SIZE,
  maxProductSize = MAX_SIZE,
  resolution = RESOLUTION,
  levels = LEVELS,
  enzymes = ENZYMES
)
```

```{r oral_l1_rflp}
data(RESTRICTION_ENZYMES)

TYPE <- 'melt'
LEVELS <- 4
MIN_LENGTH <- 15
MAX_LENGTH <- 25
MIN_SIZE <- 60
MAX_SIZE <- 200
RESOLUTION <- seq(75, 300, 15)

lapply(
  rep(sample(1:3, 3), 3),  # Random RE combinations
  function(i)
    DesignSignatures(
      dbConn,
      type = TYPE,
      enzymes = RESTRICTION_ENZYMES[i],
      minProductSize = MIN_SIZE,
      maxProductSize = MAX_SIZE,
      resolution = RESOLUTION, 
      levels = LEVELS
    )
)

# TYPE <- 'length'
# LEVELS <- 2
# MIN_SIZE <- 200
# MAX_SIZE <- 1400
# RESOLUTION <- c(
#   seq(200, 700, 3),
#   seq(705, 1000, 5),
#   seq(1010, 1400, 10)
# )
# 
# TYPE <- 'melt'
# MIN_SIZE <- 55
# MAX_SIZE <- 400
```

#### RFLP optimization by grid search

```{r grid_search}
params <- expand.grid(
  enzymes = lapply(1:3, function(x) 
    RESTRICTION_ENZYMES[
      sample(1:length(RESTRICTION_ENZYMES), 3)
    ]
  ),
  minProductSize = seq(80, 150, length.out = 3),
  maxProductSize = seq(500, 5000, length.out = 3),
  resolution = seq(1, 2, length.out = 3), 
  levels = seq(5, 20, length.out = 3)
) %>% mutate(type = rep('melt', 243))

gridsearch <- apply(params, 1, function(...) 
    list(arg = as.list(...), out = do.call(DesignSignatures, list(dbFile = dbConn, tblName = 'Seqs', identifier = 'HPV18', ...)))
) %>% bind_rows

objective <- function(x) {
  print(x)
}
```

## Anal Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
 "Human papillomavirus"[Primary Organism]
 AND viruses[filter] 
 NOT Polyamides[All Fields] 
 NOT Method[All Fields] 
 NOT Patent[All Fields]
 AND Anal[All Fields]
```

## Cutaneous Isolates
[https://www.ncbi.nlm.nih.gov/nuccore](https://www.ncbi.nlm.nih.gov/nuccore)
```{sql, eval=F}
"Human papillomavirus"[Primary Organism]
NOT Polyamides[All Fields] 
NOT Method[All Fields] 
NOT Patent[All Fields]
AND Anal[All Fields]
AND "Complete Genome"[All Fields]
```

```{r disconnect, eval=F}
tryCatch({dbDisconnect(dbconn)}, error=warning)
tryCatch({dbDisconnect(dbConn)}, error=warning)
```

